Binning Strategies and Related Loss for Large Data

### 2015-03-19 Paper.Rnw
source in binning functions and clean up code chunks for figure 3

<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
library(reshape2)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)
library(hexbin)
library(MASS)

@@

library(ggplot2)

@@

library(plyr)

@@

X$xbin <- round(X$x/binwidth,0)*binwidth
X$ybin <- round(X$y/binwidth,0)*binwidth
Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey90") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey90") + theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))

@@

X$xbin <- round(X$x/binwidth,0)*binwidth
X$ybin <- round(X$y/binwidth,0)*binwidth
Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey80") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey80")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))

@@

X$xbin <- round(X$x/binwidth,0)*binwidth
X$ybin <- round(X$y/binwidth,0)*binwidth
Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey70") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey70")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))

@@

X$xbin <- round(X$x/binwidth,0)*binwidth
X$ybin <- round(X$y/binwidth,0)*binwidth
Xsmall <- ddply(X, .(xbin, ybin), summarize, count=length(xbin))
qplot(x,y, asp=1, geom="blank") + geom_tile(binwidth=binwidth, aes(x=xbin, y=ybin, fill=count), data=Xsmall) + ggtitle(sprintf("binwidth = %.2f", binwidth)) + geom_vline(aes(xintercept=seq(min(X$xbin)-binwidth, max(X$xbin), by=binwidth)+binwidth/2), colour="grey60") + geom_hline(aes(yintercept=seq(min(X$ybin)-binwidth, max(X$ybin), by=binwidth)+binwidth/2), colour="grey60")+ theme_bw()+ theme(legend.position="bottom") +scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend", trans="log", breaks=c(1, 2, 5, 10, 20))

@@

<<spatlossdemo2,echo=F,include=T,eval=T,fig.width=12, fig.height=6, out.width='1\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Visualization of spatial loss for same data using standard, random and post-processed random binning algorithms.  Note that the total spatial loss for the standard algorithm is smaller than by random binning. The net spatial loss for the original random binned data is the total spatial loss of the post-processed assignment." >>=

RandRectBin1d <- function(xs, origin, width){
  bx1 <- origin + width/2
  bxJ <- origin + width*(floor((max(xs)-bx1)/width) + .5)
  lbs <- bx1 + width*floor((xs-bx1)/width)
  ubs <- bx1 + width*ceiling((xs-bx1)/width)
  # initially assign all values to upper bound
  binxs <- ubs
  # then use runif to reassign based on distance from 
  plower <- (ubs - xs)/width
  lowerindex <- (plower > runif(length(xs), 0, 1))
  binxs[lowerindex] <- lbs[lowerindex]
  binxs[xs < bx1] <- bx1
  binxs[xs > bxJ] <- bxJ
  return(binxs)
}
#---------------------------------------------------
n=12
param1 = 1
param2 = 1.6
set.seed(52)
x <- rbeta(n,param1, param2)*20-10
y <- rbeta(n,param1, param2)*20-10
# standard bin centers
xcent <- rep(-5,n) ; xcent[x>0] <- 5
ycent <- rep(-5,n) ; ycent[y>0] <- 5
# standard bin gridlines
temp <- data.frame(   
  x=c( -10,   0,  10, -10, -10, -10), 
  xend=c( -10,   0,  10,  10, 10, 10), 
  y=c( -10, -10 , -10 , -10, 0 , 10), 
  yend=c(10, 10 , 10, -10, 0, 10 ))

>>>>

<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
library(reshape2)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)
library(hexbin)
library(MASS)
#source in binning functions
source("BinningLossFunctions.R")

@@

@@

@@

binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
qplot(x,y, asp=1, geom="blank") +
  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
                                by=binwidth)+binwidth/2), colour="grey90") + 
  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
                                by=binwidth)+binwidth/2), colour="grey90") + 
  theme_bw()+ theme(legend.position="bottom") +
  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
                      trans="log", breaks=c(1, 2, 5, 10, 20))+
  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())

@@

binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
qplot(x,y, asp=1, geom="blank") +
  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
                                by=binwidth)+binwidth/2), colour="grey90") + 
  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
                                by=binwidth)+binwidth/2), colour="grey90") + 
  theme_bw()+ theme(legend.position="bottom") +
  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
                      trans="log", breaks=c(1, 2, 5, 10, 20))+
  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())

@@

binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
qplot(x,y, asp=1, geom="blank") +
  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
                                by=binwidth)+binwidth/2), colour="grey90") + 
  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
                                by=binwidth)+binwidth/2), colour="grey90") + 
  theme_bw()+ theme(legend.position="bottom") +
  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
                      trans="log", breaks=c(1, 2, 5, 10, 20))+
  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())

@@

binout1 <- RectBin2d(x,y,-binwidth/2,-binwidth/2,binwidth,binwidth,type="standard")[[1]]
qplot(x,y, asp=1, geom="blank") +
  geom_tile(binwidth=binwidth, aes(x=binxs, y=binys, fill=binfreq), data=binout1) +
  ggtitle(sprintf("binwidth = %.2f", binwidth)) +
  geom_vline(aes(xintercept=seq(min(binout1$binxs)-binwidth, max(binout1$binxs),
                                by=binwidth)+binwidth/2), colour="grey90") + 
  geom_hline(aes(yintercept=seq(min(binout1$binys)-binwidth, max(binout1$binys),
                                by=binwidth)+binwidth/2), colour="grey90") + 
  theme_bw()+ theme(legend.position="bottom") +
  scale_fill_gradient(low="#56B1F7", high="#132B43", guide="legend",
                      trans="log", breaks=c(1, 2, 5, 10, 20))+
  theme(panel.grid.major= element_blank(),panel.grid.minor= element_blank())

@@

<<spatlossdemo2,echo=F,include=T,eval=T,fig.width=12, fig.height=6, out.width='1\\linewidth', fig.pos='H',fig.align='center',tidy=F, cache=TRUE, fig.show='hold',fig.cap="Visualization of spatial loss for same data using standard, random and post-processed random binning algorithms.  Note that the total spatial loss for the standard algorithm is smaller than by random binning. The net spatial loss for the original random binned data is the total spatial loss of the post-processed assignment." >>= 
n=12
param1 = 1
param2 = 1.6
set.seed(52)
x <- rbeta(n,param1, param2)*20-10
y <- rbeta(n,param1, param2)*20-10
# standard bin centers
xcent <- rep(-5,n) ; xcent[x>0] <- 5
ycent <- rep(-5,n) ; ycent[y>0] <- 5
# standard bin gridlines
temp <- data.frame(   
  x=c( -10,   0,  10, -10, -10, -10), 
  xend=c( -10,   0,  10,  10, 10, 10), 
  y=c( -10, -10 , -10 , -10, 0 , 10), 
  yend=c(10, 10 , 10, -10, 0, 10 ))
